---
title: "ACF issues"
author: "Steve and Sara"
date: "October 27, 2016"
output: html_document
---

```{r libraries, include=FALSE}
library(ggplot2)
library(readr)
library(astsa)
library(fGarch)
library(forecast)
library(dplyr)

train_series <- read_csv("../data/train_data.csv", na="X") %>% 
    filter(YEAR>=2003) %>% 
    select(time=X1, ILI=X.UNWEIGHTED.ILI)

test_series <- read_csv("../data/test_data.csv") %>% 
    select(time=X1, ILI=X.UNWEIGHTED.ILI)
```

Look at the data, raw. Before trying to make it stationary.

```{r raw, echo=FALSE, fig.keep=TRUE, results='hide'}
ggplot() +
    geom_line(aes(x=train_series$time, y=train_series$ILI))
acf2(train_series$ILI,200)
```

Look at the differenced data

```{r diff, echo=FALSE, fig.keep=TRUE, results='hide'}
ggplot() +
    geom_line(aes(x=train_series$time[-1], y=diff(train_series$ILI)))
acf2(diff(train_series$ILI),200)
```

Look at the difference of the log data

```{r diff log, echo=FALSE, fig.keep=TRUE, results='hide'}
ggplot() +
    geom_line(aes(x=train_series$time[-1], y=diff(log(train_series$ILI))))
acf2(diff(log(train_series$ILI)),200)
```

Differenced log data with seasonal differencing

```{r diff diff log, echo=FALSE, fig.keep=TRUE, results='hide'}
ggplot() +
    geom_line(aes(x=1:456, y=diff(diff(log(train_series$ILI),52))))
acf2(diff(diff(log(train_series$ILI),52)),200)
```

Differenced non-log data with seasonal differencing

```{r diff diff, echo=FALSE, fig.keep=TRUE, results='hide'}
ggplot() +
    geom_line(aes(x=1:456, y=diff(diff(train_series$ILI),52)))
acf2(diff(diff(train_series$ILI),52),200)
```

These all look varying levels of stationary to us... what do the SARIMA models for each show? We would expect it to be poor for the raw data -- since it is clearly not stationary -- but the others could all work or not.

Raw data SARIMA:

```{r raw sarima, echo=FALSE, fig.keep=TRUE}
ILI_auto_arima <- auto.arima(train_series$ILI,
                             stepwise = FALSE,
                             approximation = FALSE,
                             trace=FALSE,
                             seasonal=TRUE)
arimaorder(ILI_auto_arima)
```

```{r, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(train_series$ILI, 2,0,0,0,0,0,0)
```

The p values for Ljung-Box look terrible; lets try forcing some seasonality:

```{r raw seasonal sarima, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(train_series$ILI, 2,0,0,0,1,0,52)
```

Still not good.

Differenced data SARIMA:

```{r diff sarima, echo=FALSE}
ILI_auto_arima <- auto.arima(diff(train_series$ILI),
                             stepwise = FALSE,
                             approximation = FALSE,
                             trace=FALSE)
arimaorder(ILI_auto_arima)
```

```{r, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(train_series$ILI), 2,0,0,0,0,0,0)
```

With a seasonal period of 52:

```{r diff seasonal sarima, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(train_series$ILI), 2,0,0,0,1,0,52)
```

Differenced log data SARIMA:

```{r diff log sarima, echo=FALSE}
ILI_auto_arima <- auto.arima(diff(log(train_series$ILI)),
                             stepwise = FALSE,
                             approximation = FALSE,
                             trace=FALSE)
arimaorder(ILI_auto_arima)
```

```{r, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(log(train_series$ILI)), 1,0,1,0,0,0,0)
```

With a seasonal period of 52:

```{r diff log seasonal sarima, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(log(train_series$ILI)), 1,0,1,0,1,0,52)
```

That is pretty good!

Differenced log data with seasonal differencing SARIMA:

```{r diff diff log sarima, echo=FALSE}
ILI_auto_arima <- auto.arima(diff(diff(log(train_series$ILI)),52),
                             stepwise = FALSE,
                             approximation = FALSE,
                             trace=FALSE)
arimaorder(ILI_auto_arima)
```

```{r, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(diff(log(train_series$ILI)),52), 2,0,2,0,0,0,0)
```

Somehow this pre-differenced data doesn't help much.

Differenced non-log data with seasonal differencing SARIMA:

```{r diff diff sarima, echo=FALSE}
ILI_auto_arima <- auto.arima(diff(diff(train_series$ILI),52),
                             stepwise = FALSE,
                             approximation = FALSE,
                             trace=FALSE)
arimaorder(ILI_auto_arima)
```

```{r, echo=FALSE, fig.keep=TRUE, results='hide'}
sarima(diff(diff(train_series$ILI),52), 4,0,0,0,0,0,0)
```
